\documentclass[a4paper,12pt]{article}

\setlength{\parskip}{1ex plus1pt} % threequarters
\RequirePackage[hmargin=3.5cm,vmargin=3cm]{geometry}
\usepackage{graphicx}
\pagestyle{empty}
\usepackage[osf]{mathpazo}

\begin{document}

{\raggedleft
  Matthew Pennell\\
  Institute for Bioinformatics and Evolutionary Studies\\
  University of Idaho\\
 Moscow, ID 83844, U.S.A.\\[2ex]
}

\vspace{3ex}

Dear Drs. Gibson and Sandhu,

We would like to resubmit our manuscript ``How much of the world is woody?'' (JEcol-2013-0465) 
for your re--consideration as a ``Future Directions'' article in the Journal of Ecology.
We received a rejection decision on 3 Oct. 2013 and were encouraged to resubmit our article.

We appreciate the thoroughness of the reviewers' comments and feel that our submission is much 
stronger as a result.

The most serious concerns of the reviewers can be broadly construed as falling into four categories:
1) that our definition of the woody habit was perhaps inappropriate; 2) that we failed to adequately
demonstrate how we got from the raw data to the final resuls; 3) that we did not adequately 
present the full results of the analysis; and 4) that we did not sufficiently explore the sensitivity
of the analyses to errors in the dataset and/or our coding of woody vs. herbaceous.

To address these criticisms we have substantially overhauled our submission. In addition to making changes
to the manuscript, our supplemental materials have expanded substantially. The raw data is now
included with all the steps necessary to perform the analysis. We have investigated the robustness of 
our results to various sources of bias. And we include a far more complete presentation of the results
from every subclade analyzed. This will hopefully make our results both more transparent and useful
to other resesarchers.

We also note that just prior to our resubmission, a new version of the Plant List was released, which
contained a more up--to--date list of accepted names and synonymies. We have re--done all analyses using this new 
resource; while the overall message of the paper is still the same, the actual numbers differ from
the original submission. 
 
We hope that you will consider this submission acceptable for publication in your journal and 
look forward to hearing from you.


\begin{flushright}
\vspace{2ex}
\hspace{.2\textwidth}Sincerely,\\
\hspace*{.3\textwidth}
Matthew Pennell
\end{flushright}

\end{document}
